---
title: "NFL_Betting"
author: "Kaden Kozak"
date: "2023-11-20"
output:
  pdf_document: default
  html_document: default
editor_options:
  markdown:
    wrap: 72
---

**Background:** This data set is the lines of the NFL spreads and
over/unders since 1979.The spread of the game is considered the
projected margin of victory by the favorited team and the over/under is
the total projected points to be scored in the game by both teams.

Ex: if the Vikings are playing the Packers and the spread is Vikings -3
they are projected to win the game by 3 points and they are favorites
and Green Bay is the underdog, so if the Vikings win by 7 the favorite
team covers and if they only win by one in this instance the underdog
Packers cover .

What I am trying to do with this data is find certain groups or
categories that cover the spread more often than the 50/50 likelyhood
that the line is set at. I did not look at specific teams because I feel
it would not have a good at seeing trends because teams are constantly
changing especially over a long period of time.

```{r}
library(ggplot2)
library(dplyr)
library(forcats)
library(gridExtra)
```

```{r}
library(readxl)
spreadspoke_scores <- read_excel("spreadspoke_scores.xlsx")
```

```{r}
#making the team names uniform
spreadspoke_scores[spreadspoke_scores$team_home=='Baltimore Colts',"team_home"]<-'Indianapolis Colts'
spreadspoke_scores[spreadspoke_scores$team_away=='Baltimore Colts',"team_away"]<-'Indianapolis Colts'
spreadspoke_scores[spreadspoke_scores$team_home=='Houston Oilers',"team_home"]<-'Tennessee Titans'
spreadspoke_scores[spreadspoke_scores$team_away=='Houston Oilers',"team_away"]<-'Tennessee Titans'
spreadspoke_scores[spreadspoke_scores$team_home=='Tennessee Oilers',"team_home"]<-'Tennessee Titans'
spreadspoke_scores[spreadspoke_scores$team_away=='Tennessee Oilers',"team_away"]<-'Tennessee Titans'
spreadspoke_scores[spreadspoke_scores$team_home=='Los Angeles Raiders',"team_home"]<-'Oakland Raiders'
spreadspoke_scores[spreadspoke_scores$team_away=='Los Angeles Raiders',"team_away"]<-'Oakland Raiders'
spreadspoke_scores[spreadspoke_scores$team_home=='Phoenix Cardinals',"team_home"]<-'Arizona Cardinals'
spreadspoke_scores[spreadspoke_scores$team_away=='Phoenix Cardinals',"team_away"]<-'Arizona Cardinals'
spreadspoke_scores[spreadspoke_scores$team_home=='San Diego Chargers',"team_home"]<-'Los Angeles Chargers'
spreadspoke_scores[spreadspoke_scores$team_away=='San Diego Chargers',"team_away"]<-'Los Angeles Chargers'
spreadspoke_scores[spreadspoke_scores$team_home=='St. Louis Cardinals',"team_home"]<-'Arizona Cardinals'
spreadspoke_scores[spreadspoke_scores$team_away=='St. Louis Cardinals',"team_away"]<-'Arizona Cardinals'
spreadspoke_scores[spreadspoke_scores$team_home=='St. Louis Rams',"team_home"]<-'Los Angeles Rams'
spreadspoke_scores[spreadspoke_scores$team_away=='St. Louis Rams',"team_away"]<-'Los Angeles Rams'
spreadspoke_scores[spreadspoke_scores$team_home=='Washington Commanders',"team_home"]<-'Washington Redskins'
spreadspoke_scores[spreadspoke_scores$team_away=='Washington Commanders',"team_away"]<-'Washington Redskins'
spreadspoke_scores[spreadspoke_scores$team_home=='Washington Football Team',"team_home"]<-'Washington Redskins'
spreadspoke_scores[spreadspoke_scores$team_away=='Washington Football Team',"team_away"]<-'Washington Redskins'
spreadspoke_scores[spreadspoke_scores$team_home=='Las Vegas Raiders',"team_home"]<-'Oakland Raiders'
spreadspoke_scores[spreadspoke_scores$team_away=='Las Vegas Raiders',"team_away"]<-'Oakland Raiders'
```

```{r}
spreadspoke_scores[spreadspoke_scores$team_home=='Arizona Cardinals',"team_home"]<-'ARI'
spreadspoke_scores[spreadspoke_scores$team_away=='Arizona Cardinals',"team_away"]<-'ARI'
spreadspoke_scores[spreadspoke_scores$team_home=='Atlanta Falcons',"team_home"]<-'ATL'
spreadspoke_scores[spreadspoke_scores$team_away=='Atlanta Falcons',"team_away"]<-'ATL'
spreadspoke_scores[spreadspoke_scores$team_home=='Baltimore Ravens',"team_home"]<-'BAL'
spreadspoke_scores[spreadspoke_scores$team_away=='Baltimore Ravens',"team_away"]<-'BAL'
spreadspoke_scores[spreadspoke_scores$team_home=='Buffalo Bills',"team_home"]<-'BUF'
spreadspoke_scores[spreadspoke_scores$team_away=='Buffalo Bills',"team_away"]<-'BUF'
spreadspoke_scores[spreadspoke_scores$team_home=='Carolina Panthers',"team_home"]<-'CAR'
spreadspoke_scores[spreadspoke_scores$team_away=='Carolina Panthers',"team_away"]<-'CAR'
spreadspoke_scores[spreadspoke_scores$team_home=='Chicago Bears',"team_home"]<-'CHI'
spreadspoke_scores[spreadspoke_scores$team_away=='Chicago Bears',"team_away"]<-'CHI'
spreadspoke_scores[spreadspoke_scores$team_home=='Cincinnati Bengals',"team_home"]<-'CIN'
spreadspoke_scores[spreadspoke_scores$team_away=='Cincinnati Bengals',"team_away"]<-'CIN'
spreadspoke_scores[spreadspoke_scores$team_home=='Cleveland Browns',"team_home"]<-'CLE'
spreadspoke_scores[spreadspoke_scores$team_away=='Cleveland Browns',"team_away"]<-'CLE'
spreadspoke_scores[spreadspoke_scores$team_home=='Dallas Cowboys',"team_home"]<-'DAL'
spreadspoke_scores[spreadspoke_scores$team_away=='Dallas Cowboys',"team_away"]<-'DAL'
spreadspoke_scores[spreadspoke_scores$team_home=='Denver Broncos',"team_home"]<-'DEN'
spreadspoke_scores[spreadspoke_scores$team_away=='Denver Broncos',"team_away"]<-'DEN'
spreadspoke_scores[spreadspoke_scores$team_home=='Detroit Lions',"team_home"]<-'DET'
spreadspoke_scores[spreadspoke_scores$team_away=='Detroit Lions',"team_away"]<-'DET'
spreadspoke_scores[spreadspoke_scores$team_home=='Green Bay Packers',"team_home"]<-'GB'
spreadspoke_scores[spreadspoke_scores$team_away=='Green Bay Packers',"team_away"]<-'GB'
spreadspoke_scores[spreadspoke_scores$team_home=='Houston Texans',"team_home"]<-'HOU'
spreadspoke_scores[spreadspoke_scores$team_away=='Houston Texans',"team_away"]<-'HOU'
spreadspoke_scores[spreadspoke_scores$team_home=='Indianapolis Colts',"team_home"]<-'IND'
spreadspoke_scores[spreadspoke_scores$team_away=='Indianapolis Colts',"team_away"]<-'IND'
spreadspoke_scores[spreadspoke_scores$team_home=='Jacksonville Jaguars',"team_home"]<-'JAX'
spreadspoke_scores[spreadspoke_scores$team_away=='Jacksonville Jaguars',"team_away"]<-'JAX'
spreadspoke_scores[spreadspoke_scores$team_home=='Kansas City Chiefs',"team_home"]<-'KC'
spreadspoke_scores[spreadspoke_scores$team_away=='Kansas City Chiefs',"team_away"]<-'KC'
spreadspoke_scores[spreadspoke_scores$team_home=='Los Angeles Chargers',"team_home"]<-'LAC'
spreadspoke_scores[spreadspoke_scores$team_away=='Los Angeles Chargers',"team_away"]<-'LAC'
spreadspoke_scores[spreadspoke_scores$team_home=='Los Angeles Rams',"team_home"]<-'LAR'
spreadspoke_scores[spreadspoke_scores$team_away=='Los Angeles Rams',"team_away"]<-'LAR'
spreadspoke_scores[spreadspoke_scores$team_home=='Miami Dolphins',"team_home"]<-'MIA'
spreadspoke_scores[spreadspoke_scores$team_away=='Miami Dolphins',"team_away"]<-'MIA'
spreadspoke_scores[spreadspoke_scores$team_home=='Minnesota Vikings',"team_home"]<-'MIN'
spreadspoke_scores[spreadspoke_scores$team_away=='Minnesota Vikings',"team_away"]<-'MIN'
spreadspoke_scores[spreadspoke_scores$team_home=='New England Patriots',"team_home"]<-'NE'
spreadspoke_scores[spreadspoke_scores$team_away=='New England Patriots',"team_away"]<-'NE'
spreadspoke_scores[spreadspoke_scores$team_home=='New Orleans Saints',"team_home"]<-'NO'
spreadspoke_scores[spreadspoke_scores$team_away=='New Orleans Saints',"team_away"]<-'NO'
spreadspoke_scores[spreadspoke_scores$team_home=='New York Giants',"team_home"]<-'NYG'
spreadspoke_scores[spreadspoke_scores$team_away=='New York Giants',"team_away"]<-'NYG'
spreadspoke_scores[spreadspoke_scores$team_home=='New York Jets',"team_home"]<-'NYJ'
spreadspoke_scores[spreadspoke_scores$team_away=='New York Jets',"team_away"]<-'NYJ'
spreadspoke_scores[spreadspoke_scores$team_home=='Oakland Raiders',"team_home"]<-'LVR'
spreadspoke_scores[spreadspoke_scores$team_away=='Oakland Raiders',"team_away"]<-'LVR'
spreadspoke_scores[spreadspoke_scores$team_home=='Philadelphia Eagles',"team_home"]<-'PHI'
spreadspoke_scores[spreadspoke_scores$team_away=='Philadelphia Eagles',"team_away"]<-'PHI'
spreadspoke_scores[spreadspoke_scores$team_home=='Pittsburgh Steelers',"team_home"]<-'PIT'
spreadspoke_scores[spreadspoke_scores$team_away=='Pittsburgh Steelers',"team_away"]<-'PIT'
spreadspoke_scores[spreadspoke_scores$team_home=='San Francisco 49ers',"team_home"]<-'SF'
spreadspoke_scores[spreadspoke_scores$team_away=='San Francisco 49ers',"team_away"]<-'SF'
spreadspoke_scores[spreadspoke_scores$team_home=='Seattle Seahawks',"team_home"]<-'SEA'
spreadspoke_scores[spreadspoke_scores$team_away=='Seattle Seahawks',"team_away"]<-'SEA'
spreadspoke_scores[spreadspoke_scores$team_home=='Tampa Bay Buccaneers',"team_home"]<-'TB'
spreadspoke_scores[spreadspoke_scores$team_away=='Tampa Bay Buccaneers',"team_away"]<-'TB'
spreadspoke_scores[spreadspoke_scores$team_home=='Tennessee Titans',"team_home"]<-'TEN'
spreadspoke_scores[spreadspoke_scores$team_away=='Tennessee Titans',"team_away"]<-'TEN'
spreadspoke_scores[spreadspoke_scores$team_home=='Washington Redskins',"team_home"]<-'WAS'
spreadspoke_scores[spreadspoke_scores$team_away=='Washington Redskins',"team_away"]<-'WAS'
```

```{r}
# Create new columns 'home_division' and 'away_division' and initialize them as NA
spreadspoke_scores$home_division <- NA
spreadspoke_scores$away_division <- NA


# Define divisions based on the shortened team names
afc_east <- c('BUF', 'MIA', 'NE', 'NYJ')
afc_north <- c('BAL', 'CIN', 'CLE', 'PIT')
afc_south <- c('HOU', 'IND', 'JAX', 'TEN')
afc_west <- c('DEN', 'KC', 'LAC', 'LVR')
nfc_east <- c('DAL', 'NYG', 'PHI', 'WAS')
nfc_north <- c('CHI', 'DET', 'GB', 'MIN')
nfc_south <- c('ATL', 'CAR', 'NO', 'TB')
nfc_west <- c('ARI', 'LAR', 'SF', 'SEA')

spreadspoke_scores <- spreadspoke_scores %>%
  mutate(
    home_division = case_when(
      team_home %in% afc_east ~ 'AFC East',
      team_home %in% afc_north ~ 'AFC North',
      team_home %in% afc_south ~ 'AFC South',
      team_home %in% afc_west ~ 'AFC West',
      team_home %in% nfc_east ~ 'NFC East',
      team_home %in% nfc_north ~ 'NFC North',
      team_home %in% nfc_south ~ 'NFC South',
      team_home %in% nfc_west ~ 'NFC West',
      TRUE ~ NA_character_
    ),
    away_division = case_when(
      team_away %in% afc_east ~ 'AFC East',
      team_away %in% afc_north ~ 'AFC North',
      team_away %in% afc_south ~ 'AFC South',
      team_away %in% afc_west ~ 'AFC West',
      team_away %in% nfc_east ~ 'NFC East',
      team_away %in% nfc_north ~ 'NFC North',
      team_away %in% nfc_south ~ 'NFC South',
      team_away %in% nfc_west ~ 'NFC West',
      TRUE ~ NA_character_
    )
  )


```

```{r}
#create column when divisional matchups take place
spreadspoke_scores <- spreadspoke_scores %>%
  mutate(
    matchup_type = case_when(
      home_division == away_division ~ home_division,
      TRUE ~ 'Non-Divisional'
    )
  )
```

```{r}
#spread difference results
spreadspoke_scores <- spreadspoke_scores %>%
  mutate(spread_difference = case_when(
    team_home == team_favorite_id ~ score_away - score_home,
    team_away == team_favorite_id ~ score_home - score_away,
    team_favorite_id == "PICK" ~ score_home - score_away,
    TRUE ~ NA_real_
  ))

```

```{r}
#making column to show the result of spread
spreadspoke_scores <- spreadspoke_scores %>%
  mutate(spread_results = case_when(
    spread_difference < spread_favorite ~ "Favorite",
    spread_difference > spread_favorite ~ "Underdog",
    spread_difference == spread_favorite ~ "Push",
    TRUE ~ NA_character_
  ))

```

```{r}
#making row for which team is favorited
spreadspoke_scores <- spreadspoke_scores %>%
  mutate(location_favorite = case_when(
    team_favorite_id == "PICK" ~ "No Favorite",
    team_home == team_favorite_id ~ "Home",
    team_away == team_favorite_id ~ "Away",
    TRUE ~ "None"
  ))

```

```{r}
#results of over under code
spreadspoke_scores$over_under_results <- case_when(
  spreadspoke_scores$over_under_line < spreadspoke_scores$total_points ~ "Over",
  spreadspoke_scores$over_under_line > spreadspoke_scores$total_points ~ "Under",
  TRUE ~ "PUSH"
)

```

```{r}
#making column to show the result of spread
spreadspoke_scores <- spreadspoke_scores %>%
  mutate(home_spread = 
           ifelse(team_favorite_id==team_home,spread_favorite,-1*spread_favorite)
  )

```

```{r}
#home team cover column
spreadspoke_scores <- spreadspoke_scores %>%
  mutate(home_cover = case_when(
    score_difference < home_spread  ~ "Covered",
    score_difference > home_spread  ~ "Not Covered",
    spread_results == "Push" ~ "Push",
    TRUE ~ NA_character_
  ))

```

```{r}
# Create a new variable for spread categories
spreadspoke_scores <- spreadspoke_scores %>%
  mutate(spread_category = case_when(
    spread_favorite <= -7.5 ~ "Large Spread",
    spread_favorite <= -3.5 & spread_favorite >= -7 ~ "Medium Spread",
    spread_favorite <= 0 & spread_favorite >= -3 ~ "Small Spread",
    TRUE ~ "No Spread"  # If none of the conditions match
  ))
```

```{r}
#Create new variable for over_under_category
spreadspoke_scores <- spreadspoke_scores %>%
  mutate(over_under_category = case_when(
    over_under_line <= 36.5 ~ "Low",
    between(over_under_line, 37, 47) ~ "Average",
    over_under_line >= 47.5 ~ "High",
    TRUE ~ NA_character_
  ))

```

```{r}
ggplot(spreadspoke_scores, aes(x = as.factor(spread_favorite), fill = as.factor(spread_favorite))) +
  geom_bar() +
  labs(title = "Spread Favorite Distribution",
       x = "Spread Favorite",
       y = "Frequency") +
  coord_cartesian(ylim = c(0, max(table(spreadspoke_scores$spread_favorite)))) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "none") +  # Remove legend since color represents spread_favorite
  scale_fill_viridis_d(option="plasma")  # Use the "viridis" color palette
```

The highest frequencies are at 1, 3, 3.5, and 7.Most of the spreads are
by 3 and 7 becasue those are the increments teams score at most often.

```{r}
# Filter the dataframe to remove NA values and the value 36.6
filtered_over_under_line <- spreadspoke_scores[!is.na(spreadspoke_scores$over_under_line) & spreadspoke_scores$over_under_line != 36.6,]

# Create a ggplot bar plot with color and improved aesthetics using the "viridis" palette
ggplot(filtered_over_under_line, aes(x = as.factor(over_under_line), fill = as.factor(over_under_line))) +
  geom_bar(width = 0.9) +  
  labs(title = "Over/Under Line Distribution",
       x = "Over/Under Line",
       y = "Frequency") +
  coord_cartesian(ylim = c(0, max(table(filtered_over_under_line$over_under_line)))) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1),  # Adjust angle and justification
        legend.position = "none") +  
  scale_fill_viridis_d()  
```

Over under lines have much a pretty normal shaped distrabution, but the
whole number spreads are much more popular and it has a longer right
tale.

```{r}
# Your existing code for the second two plots
season_summary_over_under <- spreadspoke_scores %>%
  group_by(schedule_season) %>%
  summarize(mean_over_under_line = mean(over_under_line, na.rm = TRUE))

trend_over_under_plot <- ggplot(season_summary_over_under, aes(x = as.factor(schedule_season), y = mean_over_under_line, group = 1)) +
  geom_line(color = "orange") +
  geom_point(color = "brown") +
  labs(title = "Trend of Over/Under Line by Season",
       x = "Season",
       y = "Mean Over/Under Line") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

season_summary_spread_favorite <- spreadspoke_scores %>%
  group_by(schedule_season) %>%
  summarize(mean_spread_favorite = mean(spread_favorite, na.rm = TRUE))

trend_spread_favorite_plot <- ggplot(season_summary_spread_favorite, aes(x = as.factor(schedule_season), y = mean_spread_favorite, group = 1)) +
  geom_line(color = "blue") +
  geom_point(color = "red") +
  labs(title = "Trend of Spread Favorite by Season",
       x = "Season",
       y = "Mean Spread Favorite") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Arrange the new plots using gridExtra in a 2x1 grid
grid.arrange(
  trend_over_under_plot,
  trend_spread_favorite_plot,
  ncol = 1
)

```

lower over/under= better defenses

closer spread= more competativeness

With the over/under there was a dip down in the early 90's but was on a
steady uptrend until the last few years it has gone down. There is not
much to see at with the spread besides the 1992 and 2009 having large
mean spreads throughout the year.

```{r}
# Assuming the two ggplot objects are stored in plot1 and plot2
spread <- ggplot(spreadspoke_scores, aes(x = "", fill = spread_results)) +
  geom_bar(width = 1) +
  coord_polar(theta = "y") +
  stat_count(aes(label = sprintf("%d\n%.1f%%", ..count.., 100 * ..count../sum(..count..))),
             position = position_stack(vjust = 0.5)) +
  geom_text(stat = "count", aes(label = paste0(sprintf("%.2f%%", prop.table(..count..) * 100))),
            position = position_stack(vjust = .5), color = "black", size = 3) +
  labs(title = "Spread Results Distribution",
       x = NULL,
       y = NULL) +
  theme_minimal() +
  scale_fill_manual(values = c("Favorite" = "green", "Underdog" = "red", "Push" = "blue"))

over_under <- ggplot(spreadspoke_scores, aes(x = "", fill = over_under_results)) +
  geom_bar(width = 1) +
  coord_polar(theta = "y") +
  stat_count(aes(label = sprintf("%d\n%.1f%%", ..count.., 100 * ..count../sum(..count..))),
             position = position_stack(vjust = 0.5)) +
  geom_text(stat = "count", aes(label = paste0(sprintf("%.2f%%", prop.table(..count..) * 100))),
            position = position_stack(vjust = .5), color = "black", size = 3) +
  labs(title = "Over Under Distribution",
       x = NULL,
       y = NULL) +
  theme_minimal() +
  scale_fill_manual(values = c("Over" = "orange", "Under" = "yellow", "PUSH" = "brown"))


# Arrange plots side by side using grid.arrange
grid.arrange(spread, over_under, ncol = 2)

```

Both the spread and the over/under are very close to 50/50. With the
under and underdog having slight edges.

```{r}
# Exclude "Push" category and focus on "Over" and "Under"
filtered_data <- spreadspoke_scores[spreadspoke_scores$over_under_results %in% c("Over", "Under"), ]

# Perform a proportion test
result <- prop.test(table(filtered_data$over_under_results))

# Display the test result
print(result)

```

The p-value (0.07597) is greater than the typical significance level of
0.05. Therefore, there isn't enough evidence to reject the null
hypothesis. The observed proportion (p-hat)is 0.4913961, which is close
to 0.5. The confidence interval also includes 0.5, supporting the idea
that the observed proportion is not significantly different from 0.5.

```{r}
# Create the first plot
spread_plot <- ggplot(spreadspoke_scores, aes(x = schedule_season, fill = spread_results)) +
  geom_bar(position = "fill", width = 0.7) +
  labs(title = "Spread Results by Schedule Season",
       x = "Schedule Season",
       y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("Favorite" = "green", "Underdog" = "red", "Push" = "blue"))

# Create the second plot
over_under_plot <- ggplot(spreadspoke_scores, aes(x = schedule_season, fill = over_under_results)) +
  geom_bar(position = "fill", width = 0.7) +
  labs(title = "Over/Under Results by Schedule Season",
       x = "Schedule Season",
       y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("Over" = "orange", "Under" = "yellow", "PUSH" = "brown"))

# Arrange the plots using gridExtra
grid.arrange(spread_plot, over_under_plot, ncol = 1)

```

All years are pretty close to 50/50 besides a couple years where at
highest it at 55% to the underdogs. Overall the under and underdogs have
very slight edges in most years.

```{r}
# Your existing code for the first plot
custom_order <- c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11", "12", "13", "14", "15", "16", "17", "18", "Wildcard", "Division", "Conference", "Superbowl")
spreadspoke_scores$schedule_week <- fct_relevel(spreadspoke_scores$schedule_week, custom_order)

spread_plot <- ggplot(spreadspoke_scores, aes(x = schedule_week, y = spread_favorite, group = 1)) +
  stat_summary(fun.y = mean, geom = "line", color = "blue") +
  stat_summary(fun.y = mean, geom = "point", color = "red", size = 2) +
  labs(title = "Average Spread Favorite Over Schedule Week",
       x = "Schedule Week",
       y = "Average Spread Favorite") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Your existing code for the second plot
over_under_plot <- ggplot(spreadspoke_scores, aes(x = schedule_week, y = over_under_line, group = 1)) +
  stat_summary(fun.y = mean, geom = "line", color = "orange") +
  stat_summary(fun.y = mean, geom = "point", color = "brown", size = 2) +
  labs(title = "Average Over/Under Line Over Schedule Week",
       x = "Schedule Week",
       y = "Average Over/Under Line") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Arrange the plots using gridExtra
grid.arrange(spread_plot, over_under_plot, ncol = 1)

```

Looking at the spread it is lowest on week one and trends to a larger
spread throughout the season, likely because it is more clear what teams
are capable of. Over/under is much more consistent throughout the year
until week 18 which is when lots of teams will sit thier starter if they
clinch playoffs.

```{r}
# Create a ggplot pie chart for 'location_favorite'
ggplot(spreadspoke_scores, aes(x = "", fill = as.factor(location_favorite))) +
  geom_bar(width = 1, color = "black") +
  coord_polar(theta = "y") +
  labs(title = "Location of the Favorite Team Distribution",
       x = NULL,
       y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_blank()) +  # Hide x-axis labels
  scale_fill_manual(values = c("Home" = "lightgreen", "Away" = "lightblue","No Favorite"="grey")) +
  geom_text(stat = "count", aes(label = paste0(sprintf("%.1f%%", prop.table(..count..) * 100))),
            position = position_stack(vjust = 0.5), color = "black", size = 3)

```

Home teams are the favorited team 66.5% of the time.

```{r}

# Create a ggplot pie chart with counts and percentages for 'home_cover'
ggplot(spreadspoke_scores, aes(x = "", fill = home_cover)) +
  geom_bar(width = 1) +
  coord_polar(theta = "y") +
  stat_count(aes(label = sprintf("%d\n%.1f%%", ..count.., 100 * ..count../sum(..count..))),
             position = position_stack(vjust = 0)) +
  geom_text(stat = "count", aes(label = paste0(sprintf("%.2f%%", prop.table(..count..) * 100))),
            position = position_stack(vjust = 0.5), color = "black", size = 6) +  # Adjust size
  labs(title = "Home Team Cover Distribution",
       x = NULL,
       y = NULL) +
  theme_minimal() +
  scale_fill_manual(values = c("Covered" = "green", "Not Covered" = "red", "Push" = "blue"))

```

With the different distribution from the graph before I thought there
might be some differences in the home teams and if they covered or not
but there was nearly no difference.

```{r}
# Subset the data to exclude Non-divisional matchups
subset_data <- spreadspoke_scores[spreadspoke_scores$matchup_type != "Non-Divisional", ]

# Create a ggplot pie chart with counts and percentages for 'home_cover'
ggplot(subset_data, aes(x = "", fill = home_cover)) +
  geom_bar(width = 1) +
  coord_polar(theta = "y") +
  stat_count(aes(label = sprintf("%d\n%.1f%%", ..count.., 100 * ..count../sum(..count..))),
             position = position_stack(vjust = 0)) +
  geom_text(stat = "count", aes(label = paste0(sprintf("%.2f%%", prop.table(..count..) * 100))),
            position = position_stack(vjust = 0.5), color = "black", size = 6) +  # Adjust size
  labs(title = "Home Team Cover Distribution (Excluding Non-Divisional)",
       x = NULL,
       y = NULL) +
  theme_minimal() +
  scale_fill_manual(values = c("Covered" = "green", "Not Covered" = "red", "Push" = "blue"))
```

In this chart I just looked at divisional opponents and surprisingly
home team covers less often, but there is not much of a difference and
not big enough to say it is significant.

```{r}
ggplot(spreadspoke_scores, aes(x = location_favorite, fill = spread_results)) +
  geom_bar(position = "fill", width = 0.7) +
  labs(title = "Spread Results by location of favorite",
       x = "Location of Favorite",
       y = "Percentage") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("Favorite" = "green", "Underdog" = "red", "Push" = "blue"))
```

There is no difference between the home and away with who covers the
spread. In no favorite Home is the favorite and Away is Underdog.

```{r}
# Your first plot
spread_plot <- ggplot(spreadspoke_scores, aes(x = spread_category, fill = spread_results)) +
  geom_bar(position = "fill", width = 0.7) +
  labs(title = "Spread Results by Spread Category",
       x = "Spread Category",
       y = "Percentage") +
  theme_minimal() +
  scale_fill_manual(values = c("Favorite" = "green", "Underdog" = "red", "Push" = "blue"))

# Assuming your data frame is named 'spreadspoke_scores'
over_under_category_filtered <- spreadspoke_scores %>%
  filter(!is.na(over_under_category) & over_under_category != "Low 36.5 and below")

# Your second plot
over_under_plot <- ggplot(over_under_category_filtered, aes(x = over_under_category, fill = over_under_results)) +
  geom_bar(position = "fill", width = 0.7) +
  labs(title = "Spread Results by Over/Under Category",
       x = "Over/Under Category",
       y = "Percentage") +
  theme_minimal() +
  scale_fill_manual(values = c("Over" = "orange", "Under" = "yellow", "PUSH" = "brown"))

# Arrange the plots using gridExtra
grid.arrange(spread_plot, over_under_plot, ncol = 1)

```

There is nearly no difference in the spread categories except that the
small spread has pushes more often. Underdogs cover slightly more with
the medium and large spreads. High over/unders the under covers slightly
more and when they are set low the over hits a little more often.

```{r}
#test to see if spread category is independent from spread results
contingency_table <- table(spreadspoke_scores$spread_results, 
                           spreadspoke_scores$spread_category)
                           

# Perform chi-squared test
chi_square_result <- chisq.test(contingency_table)

# Print the result
print(chi_square_result)

print(contingency_table)

```

H0: no association or relationship between the two variables 
Ha: Thereis significant relationship between the two variables 
Given the very small p-value we would reject the null so there likely is a significant
relationship between the distributions given.

```{r}
# Your first plot
spread_results_plot <- ggplot(spreadspoke_scores, aes(x = schedule_week, fill = spread_results)) +
  geom_bar(position = "fill", width = 0.7) +
  labs(title = "Spread Results by Schedule Week",
       x = "Schedule Week",
       y = "Percentage") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("Favorite" = "green", "Underdog" = "red", "Push" = "blue"))

# Your second plot
over_under_results_plot <- ggplot(spreadspoke_scores, aes(x = schedule_week, fill = over_under_results)) +
  geom_bar(position = "fill", width = 0.7) +
  labs(title = "Over/Under Results by Schedule Week",
       x = "Schedule Week",
       y = "Percentage") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("Over" = "orange", "Under" = "yellow", "PUSH" = "brown"))

# Arrange the plots using gridExtra
grid.arrange(spread_results_plot, over_under_results_plot, ncol = 1)

```

Most weeks are nearly all 50/50 but in weeks 15, 18, and the Superbowl
the underdog seems to have a slight edge Throughout the season

```{r}
# Create the first plot
spread_results_plot <-ggplot(spreadspoke_scores, aes(x = matchup_type, fill = spread_results)) +
  geom_bar(position = "fill", width = 0.7) +
  labs(title = "Spread Results by Matchup Type",
       x = "Matchup Type",
       y = "Percentage") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("Favorite" = "green", "Underdog" = "red", "Push" = "blue"))
# Create the second plot
over_under_results_plot <- ggplot(spreadspoke_scores, aes(x = matchup_type, fill = over_under_results)) +
  geom_bar(position = "fill", width = 0.7) +
  labs(title = "Spread Results by Schedule Week",
       x = "Schedule Week",
       y = "Percentage") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("Over" = "orange", "Under" = "yellow", "PUSH" = "brown"))

# Arrange the plots using gridExtra
grid.arrange(spread_results_plot, over_under_results_plot, ncol = 1)

```

With all of the divisions the under and the underdog is slightly favored
but nothing that is very significant.

**Conclusion:** When looking at aspects like week of the season, home
vs. away, divisional games, spread category, ect. There is no edge when
trying to look at these aspects because linesmakers likely input these
factors into their models to make these spreads and over/unders. So
trying to predict who will cover using these methods of
sorting/categorizing is still a coin flip. To make better predicitons I
am guessing you will have to look at more advanced offensive and
defensive performance metrics to get an edge on the linesmakers.
